---
title: "results"
author: "Vera Kvisgaard"
date: "2024-08-20"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


```{r}
library(ggplot2)
here::i_am("inst/simulations/ldags/results.rmd")
```



```{r}

indir <- here::here("inst/simulations/ldags/results/")
files <- list.files(indir, ".rds", full.names = TRUE)

# extract params from filenames
params_from_filenames <- function(files, 
                                  names = c("network", "init", "struct", "sample", "ess", "epf", "regular", "N", "r")) {
  tmp <- stringr::str_split(gsub("^(.*[\\/])|.rds", "", files), "_", simplify = FALSE)
  df <- data.frame(do.call(rbind, tmp), stringsAsFactors = F)
  names(df) <- names
  
  df$N <- factor(as.numeric(gsub("N", "", df$N)))
  
  levels <- unique(df$epf)
  levels <- levels[order(as.numeric(gsub("epf", "",  levels)))]
  df$epf <- factor(df$epf, levels)
  
  levels <- c("none", "tree", "ldag")
  df$struct <- factor(df$struct, levels)

  df
  
}


prettify_plot <- function(plot) {
  list(theme_minimal(),
       theme(legend.position = "bottom"))
}

plot_boxplot <- function(df, x = "N", y, fill = "struct", 
                         facets = as.formula("init+sample+ess+regular~network+epf")) {
  ggplot(df, aes(!! sym(x), !! sym(y), fill = !! sym(fill))) +
  facet_grid(facets) +
  geom_boxplot() +
  prettify_plot()
}
```

### Counts of each simulation setting
For each simulation setting, up to 30 simulation runs. When the table below shows
a lower frequency for some settings, the simulation is not complete.
At each run, the different structure learning procedures are applied to a single
data set sampled from the network. 
```{r}
include <- !names(df_par) %in% c("r")
data.frame(table(df_par[, include]))
```


# Asia, 8 nodes
The Asia network is the smallest and easiest to run with a wide range of different simulation settings, shown in the following. 

```{r}
files <- list.files(indir, pattern = "asia.*.rds", full.names = T)
imp <- lapply(files, readRDS)
# collect simulation results
tmp <- lapply(imp, "[[", "res")
len <- lengths(tmp)
stopifnot(all(len == len[1]))

# store in data frame
df <- cbind(params_from_filenames(files), do.call(rbind, tmp))


```
## True and false positive rate 
The true and false positive rate of the edges (TPR and FPR, respectively), averaged over all DAGs in the sample. The higher FPR when learning DAGs with local structure suggests spurious edges. Mitigated by increasing the edge penalty parameter `epf`.
```{r}
plot_boxplot(df, "N", "tpr", "struct")
plot_boxplot(df, "N", "fpr", "struct")
```


## Precision-recall 
The area under the precision recall curve, approximated by the average precision (`avgppv`). Based on the posterior edge probabilities (`edgep`) and the posterior 
ancestor relation probabilities (`arp`).
```{r}
plot_boxplot(df, "N", "edgep", "struct")
plot_boxplot(df, "N", "arp", "struct")

```

## Intervention distribution
Mean squared error of the point-estimates intervention distributions.
Tend to be more accurate with `tree`. Might sug
```{r}
plot_boxplot(df, "N", "mse", "struct")
```

