---
title: "Categorical BIDA simulations: Adjustment set size"
output: 
  html_document:
    number_sections: true
    toc: true
date: "2024-03-24"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, results = "hold")

```

```{r, prep}
# packages
library(foreach)
library(dplyr)
library(here)
library(ggplot2)

# paths
here::i_am("inst/simulations/output_adjset_size.Rmd")
inpath  <- here::here("../adjsets")                    # directory where adjsets are stored

# source helper-functions
files <- list.files(here::here("./inst/simulations/R"), full.names = T)
sapply(files, source) 

# indicator for wheter selected graphs/tables should be written to file
doExport <- FALSE
```

### Compute average adjustment set size in each simulation
```{r, cache=T}
compute_avg_adjset_size <- function(ps, n, aggregate = T) {
    # computes average adjustment set size for every cause-effect pair (x, y)
    stats <- matrix(list(), n, n)
    dindx <- diag(n) == 1
    isLocal <- length(ps$sets) == n
    for (x in seq_len(n)){
      for (y in seq_len(n)[-x]) {
      
        if (isLocal) {
          sets <- ps$sets[[x]]
          supp <- ps$support[[x]]
        } else {
          sets <- ps$sets[[x, y]]
          supp <- ps$support[[x, y]]
        }

        # find sets that indicate no effect
        indx <- apply(sets, 1, function(z) any(z == y, na.rm = T))
        if (all(indx)) {
          
          # full support for zero-effecteffect
          stats[[x, y]] <- c(n = 1,
                           wNonDe = 1,
                           wmean = NA,
                           max = NA,
                           wMax = NA)
        } else {
          
          wNonDe <- sum(supp[indx])           # support zero-effect
          wDe    <- supp[!indx]/(1-wNonDe)    # re-norm support positive effects
          size <- ncol(sets)-rowSums(is.na(sets[!indx, , drop = F]))
          
          stats[[x, y]] <- c(n = any(indx) + length(size),
                           wNonDe = wNonDe,
                           wmean = wDe%*%size,
                           max = max(size),
                           wMax = sum(wDe[size == max(size)]))
        }
      }
    }
      
    
    
    if (aggregate) {
      # averages over all pairs (and maximum size)
      tmp <- do.call(rbind, stats[!dindx])
      c(colMeans(tmp, na.rm = T), maxofmax = max(tmp[, 4], na.rm = T))
    } else {
      do.call(rbind, stats[!dindx])
    }
}


# list files where adjsets are stored 
files <- list.files(inpath, "^alarm|^child|^win95pts", full.names = T)


#test <- readRDS(f)
#compute_avg_adjset_size(test$pa_min, ncol(test$pa_min$sets))

# compute adjustment set size for each simulation setting
# - aggregate over each sampled graph and each cause-effect pair
ncores <- 4
cl <- parallel::makeCluster(ncores, type="SOCK", outfile = "")
doSNOW::registerDoSNOW(cl)
res <- foreach(f = files) %dopar% {
  imp <- readRDS(f)
  n <- dim(imp$pa_min$sets)[1]
  tmp <- t(vapply(imp, function(x) compute_avg_adjset_size(x, n = n, aggregate = T), numeric(6)))
  
  data.frame(adjset = rownames(tmp),
             tmp, 
             row.names = NULL)
}

```




Store results in a `data.frame`.
Note the distinction betweeen `Avg.max` and `Max`: the first is the maximum adjustment set size over pairs averaged over graphs, while the latter is the maximum over graphs and pairs.

```{r}

# collect simulation settings associated with each file
par <- stringr::str_split(files, ".+/|_|\\.", simplify = T)[, 2:4]
colnames(par) <- c("network", "N", "r")

# store results in data frame
indx <- rep(seq_along(res), sapply(res, nrow)) 
df   <-  data.frame(par[indx,], do.call(rbind, res))

# prettify data frame for plotting
# collect results in one data frame
labs <- c("No.unique" = "n",
          "Supp.zero" = "wNonDe",
          "Mean" ="wmean",
          "Avg.max" = "max",
          "Supp.max" = "wMax",
          "Max" = "maxofmax")

df <- rename(df, any_of(labs)) %>% 
      prettify_df()
head(df)
```


# Compare 
```{r}
tmp <- files[grepl("alarm_N1000", files)]
readRDS(tmp[1])
lens <- lapply(tmp, function(f) sapply(readRDS(f)$o$sets, nrow))
mean(unlist(lens)) # 12.4

```
### Maximum size of adjustment sets
```{r}
df %>%
  mutate(Max = as.factor(Max)) %>% 
  ggplot(aes(Max, fill = adjset)) +
  facet_grid(N~network) +
  geom_bar() # adds to 30*4 = 120
```


### Distribution over simulation runs
```{r}
plotter <- function(df, name, outpath = NULL) {
 df %>% 
    ggplot(aes(N, value, fill = adjset)) +
    facet_wrap(network + name ~., nrow = 1, scales = "free") +
    geom_boxplot() +
    plot_adjust_theme() +
    scale_fill_manual(values = col_scheme[["adjset"]], name = NULL) +
    xlab("sample size, N") -> plot
  plot
  if (doExport) {
    ggsave(paste0(outpath, "./inst/simulations/graphs/adjset_size_boxplot_", name, ".png"),
           width = 8, height = 4)
  }
  return(plot)
}

df %>% 
  tidyr::pivot_longer(any_of(names(labs))) %>% 
  group_by(network) %>%
  group_map( ~plotter(.x, .y, NULL), .keep = TRUE)
```


### Average over pairs and runs 
```{r}
df_tab <- df %>% 
  select(network, N, adjset, No.unique, Supp.zero, Mean, Max) %>% 
  group_by(network, N, adjset) %>% 
  arrange(network, N, adjset) %>% 
  summarise(across(any_of(names(labs)), ~mean(.x)), .groups = "drop")

df_tab 

if (doExport) {
  # write table to file
  filename <- here::here("./inst/simulations/tabs/adjsetsize_small2.tex")
  caption <- paste0(c("Comparison of the size of different adjustment sets in the sampled DAGs.","Averaged over every cause-effect pair and over 30 data sets independently sampled from each network."))
  label   <- paste0("tab::adjsetsize_small")
  
  df_tab %>%
    mutate(across(where(is.numeric), ~paste0(round(.x, 1)))) %>% 
    print_df_as_xtab(caption, label, filename)
}

```


### Check large minimal parent sets in win95pts
```{r large_sets, cache = T}
bnname <- "win95pts"
bn <- readRDS(paste0(here::here("./inst/data/"), bnname, ".rds"))
N <- 1000
files <- list.files(here::here("../adjsets"), sprintf("%s_N%s_r.+.rds", bnname, N), full.names = T)

# find maximum sized minimal parent set
maxsizes_all_pairs <- foreach (f = files) %do% {
  sets <- readRDS(f)$pa_min$sets
  dindx <- .row(dim(sets)) == .col(dim(sets))
  
  # find maximal set for each pair  
  maxsize <- array(dim = dim(sets))
  maxsize[!dindx] <- vapply(sets[!dindx], ncol, integer(1))
  maxsize
} 

# identify cause-effect pairs and simulation run
# unlist(list(list(1:3), list(4:5)))
maxsizes_all_pairs  <- matrix(unlist(maxsizes_all_pairs), ncol = length(maxsizes_all_pairs))
max(maxsizes_all_pairs, na.rm = T)
which_max <- which(maxsizes_all_pairs == max(maxsizes_all_pairs, na.rm = T), arr.ind = T)
n <- sqrt(nrow(maxsizes_all_pairs))
x <- (which_max[, 1]-1)%%n +1
y <- (which_max[, 1]-1)%/%n +1
r <- which_max[, 2]
cbind(r = which_max[], x, y)

```

```{r plot minimal parent sets, cache = T}

for (i in seq_len(nrow(which_max))) {
  # find DAGs where minimal parent set is of maximum size
  tmp  <- readRDS(here::here(sprintf("../partitionMCMC/%s_N%s_r%02.0f.rds", bnname, N, r[i])))
  dags <- lapply(tmp$traceadd$incidence, as.matrix)
  npar <- vapply(dags, function(dag) sum(dag[, x[i]] == 1), integer(1))
  dags_with_maxpar <- which(npar == max(npar))
  
  # first DAG with maximum-sized parent sets
  tmp <- lapply(dags[dags_with_maxpar],
                function(dag) bida:::adjsets_from_dag(c("pa_min"), dag, xvars = x[i], yvars = y[i])$pa_min)
  z <- tmp[[which.max(lengths(tmp))]]
  g <- dags_with_maxpar[which.max(lengths(tmp))]
  
  fillcolor = setNames(c("red", "blue", rep("black", length(z))), names(bn)[c(x[i], y[i], z)])
  Rgraphviz::plot(as(dags[[g]], "graphNEL"),
                  nodeAttrs = list(fillcolor = fillcolor),
                  main = paste0("Minimal parent set, x = ", x[i], ", y = ", y[i], ", r = ", r[i]))
}
```

