---
title: "Squared errors"
output: 
  html_document:
    number_sections: true
    toc: true
date: "2024-01-16"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message=FALSE, warning=FALSE, results = "hold")
here::i_am("inst/simulations/output_sqerr.Rmd")
```


## Set params
```{r}
selected_networks <- c("Child (n=20)", "Alarm (n=37)", "Win95pts (n=76)")
selected_methods <- c("BIDA+o-min", "IDA+o", "IDA+pa")
doExport <- T
```

## Load functions and import data
```{r message=FALSE, warning=FALSE}
## packages -----
library(bida)
library(doSNOW)
library(dplyr)
library(ggplot2)
library(here)
library(foreach)



files <- list.files(here::here("./inst/simulations/R"), full.names = T)
sapply(files, "source")

files <- list.files(here("../simres"), "*sqerr.rds", full.names = T)
#readRDS(files[1])

ncores <- 4
cl <- parallel::makeCluster(ncores, type="SOCK", outfile = "")
doSNOW::registerDoSNOW(cl)
imp <- foreach(f = files) %dopar% readRDS(f)

```

Store in data frame.
```{r}
# collect simulation settings from filenames
par <- stringr::str_split(files, ".+/|_|.rds", simplify = T)[, 2:5]
colnames(par) <- c("network", "N", "r", "slearn")

# store in data frame
indx <- rep.int(seq_along(imp), lengths(imp))
df <- data.frame(par[indx, ], reshape2::melt(imp))
df$L1 <- NULL


df_all <- prettify_df(df) 
   # df <- mutate(df, procedure = case_when(slearn == "MCMCPC" & method == "value" ~ paste0("BIDA+", adjset),
   #                                         slearn == "MCMCPC" & method == "rank" ~ paste0("BIDA-R+", adjset),
   #                                         slearn == "MCMCPC" & method == "arp" ~ paste0("ARP+", adjset),
   #                                         slearn == "unadjusted" ~ paste0(adjset, "."),
   #                                         slearn == "PC05" ~ paste0("IDA+", adjset)))
head(df_all)
```



## Number of simulation results
```{r}
# count rows within each simulation setting
df_all %>%
  group_by(network, slearn, N, adjset) %>%
  arrange(network, slearn, N, adjset) %>% 
 # distinct(r) %>%
  count() %>%
  tidyr::pivot_wider(names_from = "network", values_from = "n") %>% 
  print(n = 100)
```


# Overall MSE
Aggregate squared errors over true descendant and non-descendant. 
First compute number of params from the true network, then the (weighted) average over descendant status.
```{r, cache = TRUE}

bnnames <- unique(par[, "network"])
counts <- list()
for (bnname in bnnames) {
  bn <- readRDS(here::here("./inst/data/", paste0(bnname, ".rds")))
  nlev <- vapply(bn, function(x) dim(x$prob)[1], integer(1))
  dmat <- bida:::descendants(bn)

  # sum number of pairs and number of intervention-prob-table params by descendant
  counts[[bnname]] <- rowsum(cbind(npair = 1, npar = c(outer(nlev, nlev))), c(dmat))
}


df_counts <- data.frame(network = rep(bnnames, each = 2), 
                  descendant = rep(c(FALSE, TRUE), length(bnnames)),
                  do.call(rbind, counts)) %>% 
             prettify_df()

# overall mse - sum over descendant status, divide by number of params
df <- df_all
indx <- names(df) %in% c("descendant", "value")
df_mse <- aggregate(df["value"], df[!indx], "sum") %>% 
          left_join(aggregate(df_counts[, -c(1:2)], df_counts["network"], "sum"), by = "network") %>% 
          mutate(value = 1000*value/ifelse(variable == "pdo", npar, npair))

```


## Box-plots
Compute summarise statistics for box-plots.
```{r}
df_box <- df_mse %>% 
  filter(N != 10000) %>% 
  group_by(network, slearn, adjset, N, variable) %>% 
  summarise_box_plot() %>% 
  ungroup()
head(df_box)
```


### Compare adjustment sets
#### Selected networks
To limit y-axis, plot for each network separately, and combine plots using 'ggpubr::ggarrange'
```{r}

for (v in c("pdo", "jsd")) {
  for (s in c("MCMCPC", "KNOWN")) {
    
    plots <- list()
    label <- ifelse(s == "MCMCPC", "unknown", "known")
    
    for (bnname in selected_networks) {
      tmp <- df_box %>% 
              filter(N != 10000) %>% 
              filter(network == bnname) %>% 
              filter(slearn == s) %>% 
              filter(variable == v)
                
      if (nrow(tmp) == 0) next
      ylim <- c(0, max(tmp$ymax)*2/3)
        
      plot <- ggplot(tmp, aes(N, color = adjset, fill = adjset, group = group)) +
                geom_boxplot_no_points(width = .75) +
                plot_adjust_theme(axis.text.x = element_text(), ylab = "", xlab = "sample size", col_scheme[["adjset"]]) +
                facet_wrap(.~network, nrow = 1, scales = "free_y") +
                coord_cartesian(ylim = ylim)
        
      plots[[bnname]] <- plot
    }
    
    # combine plotw for each level of slearn
    plot <- ggpubr::ggarrange(plotlist = plots, nrow = 1,  common.legend = T, legend = "bottom")
    plot <- ggpubr::annotate_figure(plot,
                                    right = paste(rep.int(c(" ", label, " "), c(0, 1, 10)), collapse = ""),
                                    left = paste(rep.int(c(" ", "MSE * 1000", " "), c(15, 1, 0)), collapse = ""))
    
    # print
    print(ggpubr::annotate_figure(plot, top = paste("Variable: ", v, "G is:", label)))
    
    # export
    if (doExport) {
      filename <- paste("sqerr_box_plot_adjust_3", v, label, ".png", sep = "_")
      ggsave(here::here("inst/simulations/graphs/", filename), plot, height = 3.5, width = 8)
    }
    
  }
}
```




#### All networks
```{r, out.width = "100%"}
for (v in c("jsd")) {
  plot <- df_box %>%
    filter(variable == v) %>%
    filter(slearn %in%  c("KNOWN", "MCMCPC")) %>%
    mutate(slearn = ifelse(slearn == "MCMCPC", "unknown", "known")) %>%
  ggplot(aes(N, color = adjset, fill = adjset, group = group)) +
      geom_boxplot_no_points(width = .75) +
      plot_adjust_theme(axis.text.x = element_text(angle = 90),
                        ylab = "", xlab = "", col_scheme[["adjset"]]) +
      facet_wrap(slearn~network, nrow = 2, scales = "free") +
      scale_y_continuous(limits = c(0, NA))

  # print
  print(plot + ggtitle(sprintf("Variable: %s", v)))

  # export
  if (doExport) {
     filename <- paste("sqerr_box_plot_adjust_all_", v , ".png", sep = "_")
     ggsave(file = here("./inst/simulations/graphs", filename), plot = plot, height = 4, width = 6)
     }
}




```
#### Relative MSE

```{r}
df_mse %>% 
  filter(variable == "jsd") %>% 
  filter(slearn %in% c("KNOWN", "MCMCPC")) %>% 
  tidyr::pivot_wider(names_from = "adjset") %>% 
  mutate(slearn = ifelse(slearn == "MCMCPC", "unknown", "known"), 
         across(c("pa", "o", "pa-min", "o-min"), ~.x/`o-min`)) %>% 
  select(-`o-min`) %>% 
  tidyr::pivot_longer(c("pa","o", "pa-min"), names_to = "adjset") %>% 
  mutate(adjset = factor(adjset, c("pa", "pa-min", "o"))) %>% 
  group_by(network, slearn, adjset, N, variable) %>% 
  summarise_box_plot() %>% #ungroup() %>% distinct(slearn)
    ggplot(aes(N, color = adjset, fill = adjset, group = group)) +
        geom_abline(slope = 0, intercept = 1, linewidth = .25) +
        geom_boxplot_no_points(width = .75) +
        plot_adjust_theme(axis.text.x = element_text(angle = 90), 
                          ylab = "", xlab = "", col_scheme[["adjset"]]) +
        facet_wrap(slearn~network, nrow = 2, scales = "free_y") +
        coord_cartesian(ylim = c(0, 3))
```



### Compare BIDA and IDA
```{r}
# filter data
df <- df_box %>% 
  filter(N != 10000) %>% 
  filter(slearn %in% c("unadjusted", "MCMCPC", "PC05")) %>% 
  filter(!(slearn == "MCMCPC" & !adjset %in% c("pa", "o-min"))) %>% 
  mutate(procedure = case_when(slearn == "MCMCPC" ~ paste0("BIDA+", adjset),
                               slearn == "unadjusted" ~ adjset,
                               slearn == "PC05" ~ paste0("IDA+", adjset))) %>% 
  mutate(procedure = factor(procedure, c("cond.", "marg.", "IDA+pa", "IDA+o",  "BIDA+pa", "BIDA+o-min")))
# check 
distinct(ungroup(df), slearn, adjset, variable, procedure)
```

#### Selected networks
```{r}
plot <- df %>% 
  filter(variable == "jsd") %>% 
  filter(network %in% selected_networks) %>% 
  filter(procedure %in% selected_methods) %>% 
  ggplot(aes(N, color = procedure, fill = procedure, group = interaction(N, procedure))) +
      geom_boxplot_no_points(width = .6) +
      plot_adjust_theme(axis.text.x = element_text(), ylab = "MSE * 1000", xlab = "sample size", col_scheme[["method"]]) +
      facet_wrap(network~., scales = "free") 

print(plot)
if(doExport) {
  filename <- paste0("sqerr_box_plot_jsd_3.png")
  ggsave(file = here("./inst/simulations/graphs", filename), plot = plot, height = 4, width = 8)
}

```

#### All networks
```{r}

plot <- df %>% 
  filter(variable == "jsd") %>% 
  ggplot(aes(N, color = procedure, fill = procedure, group = interaction(N, procedure))) +
      geom_boxplot_no_points(width = .6) +
      plot_adjust_theme(axis.text.x = element_text(), ylab = "MSE * 1000", xlab = "sample size", col_scheme[["method"]]) +
      facet_wrap(network~., scales = "free") 

print(plot)
if(doExport) {
  filename <- paste0("sqerr_box_plot_jsd_all.png")
  ggsave(file = here("./inst/simulations/graphs", filename), plot = plot, height = 9, width = 7)
}

```

### Intervention probabilities
```{r}
# filter data
plot <- df %>% 
  filter(variable == "pdo") %>% 
  ggplot(aes(N, color = procedure, fill = procedure, group = interaction(N, procedure))) +
      geom_boxplot_no_points(width = .6) +
      plot_adjust_theme(axis.text.x = element_text(), ylab = "MSE * 1000", xlab = "sample size", col_scheme[["method"]]) +
      facet_wrap(network~., scales = "free") 

print(plot)
if(doExport) {
  filename <- paste0("sqerr_box_plot_pdo_all.png")
  ggsave(file = here("./inst/simulations/graphs", filename), plot = plot, height = 9, width = 7)
}

```



## Scatter plots
```{r}
plot_scatter <- function(df, v, x, y, facets = formula("adjset~network")) {
  vv <- sym(v)
  xx <- sym(x)
  yy <- sym(y)
  
  df %>% 
    filter(!! vv %in% c(x, y)) %>% 
    tidyr::pivot_wider(names_from = v) %>% 
    ggplot(aes(!! xx, !! yy, color = N)) +
    facet_grid(facets, scales = "free") + 
    geom_abline(slope = 1, intercept = 0) +
    geom_point() -> plot
  
  return(plot)
}
```

### Compare squared errors when G is known and unknwon
```{r, results = "hide", fig.keep = "all"}
group_by(df_mse, estimator, ace) %>% 
  group_split() %>% 
  lapply(function(x) plot_scatter(x, v = "slearn", x = "G", y = "MCMCPC", formula("adjset~estimator+ace+network")))

```

### Posterior mean JSD vs JSD of posterior mean CPTs
Compare the point-estimates obtained as the posterior mean (smpl_jsd) and those obtained as the JSD of the posterior mean of the intervention probabilities (cond_jsd)

```{r, results = "hide", fig.keep = "all"}
df_mse %>% 
  filter(slearn %in% c("G", "MCMCPC", "exact")) %>% 
  filter(ace == "jsd") %>% 
  group_by(slearn) %>%
  group_split() %>% 
  lapply(function(x) plot_scatter(x, v = "estimator", x = "cond", y = "smpl", formula("adjset~slearn+network")))
```

### Minimal vs maximal adjustment set 
```{r, results = "hide", fig.keep = "all"}
df_mse %>% 
  filter(slearn %in% c("G", "MCMCPC")) %>%
  filter(adjset %in% c("pa", "pa-min", "o", "o-min")) %>% 
  mutate(minimal = ifelse(grepl("min", adjset), "minimal", "full"), 
         adjset = ifelse(grepl("pa", adjset), "pa", "o")) %>% 
  group_by(slearn, estimator) %>%
  group_split() %>% 
  lapply(function(x) plot_scatter(x, v = "minimal", x = "minimal", y = "full", formula("adjset~estimator+slearn+network")))

```

# MSE by true descendant status
Aggregate squared errors over true descendant only
```{r, cache = TRUE}
df_mse <- df_all %>% 
  filter(desc == TRUE) %>% 
  group_by(network, slearn, N, r, measure, adjset, estimator, ace, desc) %>% 
  summarize(value = 1000*sum(value)/sum(n), .groups = "keep")
head(df_mse)

```

## Box-plots
```{r}
df_mse %>% 
  # keep right learner-adjset pairs and learner-estimator-pair
  filter(slearn %in% c("-", "MCMCPC", "PC05")) %>% 
  filter(ace == "jsd",
         !(slearn == "MCMCPC" & !estimator == "smpl"),
         !(slearn == "MCMCPC" & !adjset %in% c("pa", "o-min"))) %>% 
  left_join(lookup_method, by = c("slearn", "adjset", "estimator")) %>% 
  group_by(network, slearn, adjset, estimator, ace, N, method) %>% 
  summarise_box_plot() %>% 
  arrange(method) %>% 
      ggplot(aes(N, color = method, fill = method, group = group)) +
        geom_boxplot_no_points() +
        plot_adjust_theme(axis.text.x = element_text(), ylab = "MSE * 1000", xlab = "sample size", col_scheme[["method"]]) +
        facet_wrap(network~., scales = "free") 


  
```

## Compare child network to samples 
```{r}
bnname <- "child"
N <- 1000
r <- 1

filepath <- here::here("./inst/data/", paste0(bnname, ".rds"))
bn <- readRDS(filepath)
dag <- bnlearn::amat(bn)
nlev <- vapply(bn, function(x) dim(x$prob)[1], integer(1))
dindx <- .row(dim(dag)) == .col(dim(dag))
truth <- compute_ground_truth(bn)
set.seed(r+N)
data <- bida:::sample_data_from_bn(bn, N)
ps   <- parent_support_from_dags(list(dag))
bp   <- bida:::bida(ps, data, "categorical", params = list(nlev = nlev, ess = 1))
set.seed(007)
indx <- vapply(bp, function(x) is.null(x) || x$zerosupp == 1, logical(1))
smpl <- vapply(bp[!indx], posterior_sample, ace_funs =list(jsd = bida:::avg_jsd_array),n = 10**3, numeric(10**3))
tau  <- colMeans(smpl)
```
